---
title: "Basic R concepts"
output: html_notebook
---

## BASIC R CONCEPTS


### Variables 

You can consider it as a placeholder for any object. We use the assignment 
operator '<-' to assign the value. The name of the variable consists of letters, 
numbers, dots or underline characters and is case sensitive

```{r}

x <- 5
x
```


### Operators in R

Arithmetic operators

```{r}

3 + 3  #Addition
3 - 2  #Subtraction
3 * 2  #Multiplication
6 / 2  #Division
3 ^ 2  #Exponentiation
10 %% 3  #Modulus 
10 %/% 3  #Integer division

```

Mathematical functions

```{r}
abs(-5)  #Absolute 
log(2) #Natural logarithm
log(2, base = 10 ) #Logarithm
exp(5) #Exponential
factorial(5) #factorial

```

Special numbers

- infinity: typically presents a number so big or so small that cannot be 
handled by your PC
- NaN ("Not a Number") represents a value that cannot be represented by a real 
value. Typically you can come across if output of any calculation does not make
any mathematical sense 
- NA ("Not Available") presents non available or simply missing values. You will
find these in most of the datasets you will analyze

```{r}

#Infinity
1 / 0 #Positive infinity
-1 / 0 #Negative infinity
Inf + 5 #Operation on Inf

#Undefined
Inf / Inf #NaN (Not a Number)

#Missing value
NA 

```

### Logical operators

The logical operators return TRUE or FALSE, depending on the relationship between
the parameters.

```{r}
#Logical operators
5 > 2  #greater than
5 >= 2 #greater than equal to 
5 < 2  #less than
5 <= 2  #less than equal to
5 == 2 #exactly equal to
5 != 2 #not equal to
!(TRUE) #logical NOT operator
TRUE | FALSE #logical OR operator
TRUE & FALSE #logical AND Operator
```

### Data structures 

These are the different ways that data are stored and organized in the memory. 
Essentially is a collection of data elements grouped under one name. In R these 
are:

- Vector
- Factor
- List
- Dataframe 
- Matrix 
- Array 

The two most important aspects of a data structure are:

1) What kind of items does it includes?  - Similar or Dissimilar

When items are similar then the data structure is homogeneous. Atomic 
vector, matrix and array are such. 
When items are dissimilar then the data structure is heterogeneous. List and 
dataframe are such. 

2) How these items are arranged? - 1D, 2D or multi-dimensional  

Atomic vector and list are 1D data structures. 
Matrix and Dataframes are 2D data structures. 
Arrays can be multi-dimensional. 

The basic classes of objects in R are the following:
- Character e.g. "a", "B"
- Numeric e.g. 4.1, 5.5
- Integer e.g. 5, 6
- Logical  True, False


### Single dimension data structures in R

- Atomic vector
- Factor
- List

Atomic vectors are known as vectors. They are homogeneous data structures

```{r}
# Character
name <- c("maria", "george", "john")
name

# Numeric
weight <- c(65.4, 81.3, 80.5)
weight

# Integer
height <- c(175L, 183L, 185L)
height

# Logical
work <- c(TRUE, FALSE, TRUE)
work

```


Factor is a special case of vector used to store categorical values. Values are
stored as integers and each integer is assigned to a factor level. 

```{r}
# using factor
gender <- factor(c("Female", "Male", "Male"))
gender

as.numeric(gender)
```


List is a heterogeneous data structure, so that means it can contain items of 
different classes. It is also single dimensional.

```{r}
# Create an unnamed list
person1 <- list("maria", 56.4, 175L, TRUE)
person1


# Create an named list
person1 <- list(name = "maria",
                weight = 56.4, 
                height = 175L, 
                work = TRUE)
person1

```


### Higher dimensional data structures 

- Dataframe 
- Matrix 
- Array 

Dataframe is the most important data structure in R. It is heterogeneous, so it
can contain elements of different classes. It is 2-dimensional and you can
think of it as an excel spreadsheet. Each column is a different field and data
are stored in different rows. 

```{r}
# Create a dataframe using vectors
persons <- data.frame(name = name,
                     weight = weight,
                     height = height,
                     work = work)

persons
```

Matrix is similar to dataframes as it is 2-dimensional. Unlike dataframe is a
homogeneous data structure, so it contains elements of the same class only. 
Typically used to store numeric data. 

```{r}
weights <- matrix(c(65.4, 70.8, 65.7, 78.9, 82.6, 85.6, 90.8, 60.8),
                  ncol = 2,
                  nrow = 4)
weights
```

Array is also a homogeneous data structure, but it can be n-dimensional. 

```{r}
# Create two vectors of different lengths.
v1 <- c(5, 9, 3)
v2 <- c(10, 11, 12, 13, 14, 15)
column_names <- c("Col1","Col2","Col3")
row_names <- c("Row1","Row2","Row3")
matrix_names <- c("Matrix1","Matrix2")

# Take these vectors as input to the array.
result <- array(c(v1,v2),dim = c(3,3,2),dimnames = list(row_names,column_names,
   matrix_names))

result

```


### Functions 
A function is a set of statements organized together to perform a specific task. 
R has a large number of in-built functions and the user can create their own 
functions.


Built-in Functions
Simple examples of in-built functions are mean(), max(), sum(x) etc. 
They are directly called by user written programs.

```{r}
# Find the mean of numbers from 1 to 10
mean(1:10)

# Find the sum of numbers from 1 to 10
sum(1:10)

# Find the maximum number from a vector
v1 <- c(2, 5, 1, 12, 5, 8, 9, 11)
max(v1)

```


User-defined Functions

We can create user-defined functions in R. They are specific to what a user 
wants and once created they can be used like the built-in functions. 
Below is an example of how a function is created and used.

```{r}
add_2_numbers <- function(a, b) {
  print(a + b)
}

add_2_numbers(3, 6)

```

### Getting help in R

If you type the question mark symbol (?) before a function, then the documentation
for this function will appear in the help tab. 

```{r}
?mean()
```

### Load a library

This is how you can load a library in your global environment.  

```{r}
library(dplyr)
```
