---
title: "Introduction to tidyverse"
output: html_notebook
---

## BASIC TIDYVERSE

### Load libraries

How to load tidyverse libraries in R. It is the easiest way to load all core 
tidyverse packages using one command in the environment (ggplot2, purr, tibble,
dplyr, tidyr, stringr, readr, forcats)

```{r}
library(tidyverse)
```


### Pipe operator

This is an example of filtering observations from a dataframe without using the 
pipe operator

```{r}
filter(iris, Species == "setosa")
```

The pipe operator %>% basically chaining arguments together in a clear way. 

```{r}
iris %>% 
  filter(Species == "setosa")
```

Advantages:  
- You'll avoid nested function calls  
- You'll minimize the need for local variables and function definitions  
- Easy to add steps anywhere in the sequence of operations.  

As you're getting more familiar with tidyverse you will understand the power of 
this concept. 

## DATA WRANGLING

Main Libraries: dplyr, tibble, stringr, readr, forcats

### Filtering observations

Use filter() to subset observations from multiple variables.

```{r}
iris %>% 
  filter(Species == "setosa" & Sepal.Length > 4.5)
```


### Arranging observations

arrange() sorts a dataframe based on one variable

```{r}
iris %>% 
  arrange(Sepal.Length)
  
```

Use arrange() to sort a table based on two or more variables

```{r}
iris %>% 
  arrange(Sepal.Length, Petal.Length)
  
```

### Arranging observations in descending order

```{r}
iris %>% 
  arrange(desc(Sepal.Length))

```


### Change an existing variable with mutate()

mutate() adds or alter a variable

```{r}
iris %>% 
  mutate(Sepal.Length = Sepal.Length * 2)
```


### Use mutate() to add a new column

Add a new variable in the dataframe

```{r}
iris %>% 
  mutate(Sepal.Length_2 = Sepal.Length * 2)
```


### Combine Filtering, Arranging & Mutating

```{r}
iris %>% 
  filter(Species == "setosa" & Sepal.Length < 4.6) %>% 
  mutate(Sepal.Length2 = Sepal.Length * 2) %>% 
  arrange(desc(Sepal.Length2))

```



## TIDYING DATA 

Main Libraries: tidyr, tibble

Tidy datasets are easy to manipulate, model, visualize and have a specific structure.

### What is tidy data?

A set of data in a specific form:  

- Each variable is a column
- Each observation is a row  
- Each value is a cell

This is an example of a messy dataset

```{r}
datasets::WorldPhones
```

Here each column presents the number of phones in a specific area. Each row 
presents the number of phones per specific year. It is messy & very difficult to
handle e.g. analyze, model...


This is an example of a tidy dataset

```{r}
iris
```


### Gather 

Essentially it takes a "wide" dataframe with a lot of columns and transforms it 
to "long". It takes the following arguments:

1. the key, where you create the group variable
2. the value, which is the value variable (usually numeric)
3. The columns you want to group

Use gather() to make a tidy dataframe 

```{r}
WorldPhones %>% 
  as.data.frame() %>%
  rownames_to_column("year") %>% 
  gather(key = "Area", value = "Phones", N.Amer:Mid.Amer)

```


### Spread

It converts a long dataframe into a wide dataframe. In practice it reverses the
gather function, so it does the opposite of making tidy dataset. Sometimes it is
used as an intermediate step in data transformation or when creating tables 
for reports or presentations

```{r}
long_dataset <- 
  datasets::WorldPhones %>% 
  as.data.frame() %>%
  rownames_to_column("year") %>% 
  # gather(N.Amer, Europe, Asia, S.Amer, Oceania, Africa, Mid.Amer, key = "Area", value = "Phones")
  gather(key = "Area", value = "Phones", N.Amer:Mid.Amer)

long_dataset %>% 
  spread(key = "Area", value = "Phones")
```

### Separate

separate() takes string values in a single column and splits into multiple 
columns. The following arguments are used:

1. The name of the column to separate
2. The vector of new column names to separate into
3. The character to separate the values by 

```{r}
new <- data.frame(
  First = c("A_1", "A_2", "B_2", "B_3", "C_1"),
  Second = c(4, 5, 5, 6, 7)
)

new %>% 
  separate(First, into = c("main", "secondary"), sep = "_")
```



## SUMMARISING & GROUPING

Main Libraries: dplyr

### Summarize single variable

We use summarise() to produce summaries of one or more variables. A new 
dataframe is created that consists of the specified summary.

```{r}
iris %>% 
  summarise(Avg_Sepal.Length = mean(Sepal.Length))

```


### Summarize multiple variables

Use it to create more transformations. You can do as many transformations as you 
want, just separate these by commas. 

```{r}
iris %>% 
  summarise(
    Avg_Sepal.Length = mean(Sepal.Length),
    Sum_Sepal.Width = sum(Sepal.Width),
    Median_Sepal.Width = median(Sepal.Width)
    )

```


Use summarise_all() to summarise all variables. It can apply a single function to all
variables of a dataframe. If the function doesn't apply to a variable, it will 
produce a missing value.

```{r}
iris %>% 
  summarise_all(mean)

```

Use summarise_at() to summarise specific variables by applying multiple functions.

```{r}
iris %>% 
  summarise_at(vars(Sepal.Length, Sepal.Width), funs(mean, median))
```


### Group by & summarise

Pass the variable name into group_by to create an individual group for each value 
in the column. It creates a dataframe for each group and then executes the 
chosen functions. At the end it reassembles the individual output.
group_by() in conjunction with summarise() is one of the most important concepts
in tidyverse.

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(Avg_Sepal.Length = mean(Sepal.Length))
```

## VISUALISE DATA

Library: ggplot2

We'll explore the power of ggplot, which is probably the most powerful library
for visualization in any programming language. There are significant packages 
in R and other languages (e.g. matplotlib, seaborn in Python) but nothing can
compare to ggplot

There are 3 main elements in the ggplot syntax:

- 1. ggplot() function called to initiate the plots 

Initiate a plot by passing the dataframe as the first argument of the ggplot()
or use the pipe operator %>% 

ggplot(dataframe, ....) or dataframe %>% ggplot()

- 2. aes() aesthetic layer used for representing the data 

It defines the mappings to columns in the dataframe

ggplot(dataframe, aes(x = x_var)) or dataframe %>% ggplot(aes(x = x_var))

- 3. You can add multiple layers to make a plot look the way desired
Use the ‘+’ operator to add layers in the plot. The layers define what and 
how the plot gets rendered. The plot must have at least one layer in order to display

dataframe %>% 
  ggplot(aes(x = x_var)) +
  geom_histogram()


### Single Dimension plots

These are plots using a single variable. 

E.g. geom_density(), geom_dotplot(), 
geom_freqpoly(), geom_histogram(), geom_bar()


```{r}
iris %>% 
  ggplot(aes(Sepal.Length)) +
  geom_histogram()

```

You can use a lot of arguments in the geom_ layer functions. e.g.

```{r}
iris %>% 
ggplot(aes(Sepal.Length)) +
  geom_histogram(bins = 10, color = "red", fill = "orange")

```

It is easy to change the plot. e.g. below there is a density plot. We only need
to change the last line function name

```{r}
iris %>% 
ggplot(aes(Sepal.Length)) +
  geom_density()

```


### 2D plots

Create plots with 2 variables in order to show their relationship. We add an 
additional variable to the aes() function. There are a lot of geom_ functions 
but the most common are geom_point(), geom_col(), depending on the variables
data type. 

Below there is a scatter plot

```{r}
iris %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point()

```

Below there is a bar plot. Here we perform some data transformations prior to the 
plotting. 

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(Sepal.Width = mean(Sepal.Width)) %>% 
  ggplot(aes(Species, Sepal.Width)) +
  geom_col()

```


### Add layers & elements to the plot

Scatter plot with smoothed line (local regression line)

```{r}
iris %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point() + 
  geom_smooth()

```

Add a title to the previous plot

```{r}
iris %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  geom_smooth() +
  labs(
  title = "Scatter plot with smooth line for Sepal.Width and Sepal.Length"
)

```

Add a specific theme on the plot 

```{r}
iris %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  geom_smooth() +
  labs(
  title = "Scatter plot with smooth line for Sepal.Width and Sepal.Length"
) +
  theme_bw()

```

## COMBINE TIDYVERSE PACKAGES

Here we combine tidying, wrangling, grouping, summarizing and plotting data.

```{r}
WorldPhones %>% 
  
  # Make a tidy dataset
  as.data.frame() %>%
  rownames_to_column("year") %>% 
  gather(key = "Area", value = "Phones", N.Amer:Mid.Amer) %>% 
  filter(year > 1951) %>% 
  
  # Create the growth rate variable
  group_by(Area) %>% 
  mutate(previous_value = lag(Phones)) %>%
  mutate(growth_rate = (Phones-previous_value)/Phones) %>% 
  
  # Summarise the dataset
  group_by(Area) %>% 
  summarise(Avg_growth = mean(growth_rate, na.rm = TRUE)) %>% 
  
  # Create the plot
  ggplot(aes(reorder(Area, -Avg_growth), Avg_growth)) +
  geom_col() +
  labs(
    title = "Average growth rate of phones per area",
    subtitle = "Between 1956 & 1961",
    y = "Average growth rate (%)",
    x = "Area"
  ) +
  scale_y_continuous(labels=scales::percent) +
  theme_bw()

```

## MODELLING 

### Linear regression model residuals

Plot the residuals of a linear regression model 

```{r}
# Modelr is a library that works efficient with other tidyverse packages
library(modelr)

# Create a simple linear regression model
lm1 <- lm(Sepal.Length ~ Sepal.Width, data = iris)

iris %>% 
  add_residuals(lm1, "lresid") %>% 
  ggplot(aes(Sepal.Width, lresid)) +
  geom_hex(bins = 30)

```

### Multiple models

Fit multiple models in a grouped dataframe. The target here is after developing
a linear model between Sepal.Length & Sepal.Width for each Species, report the 
p value of the Sepal.Width 

```{r}
# Summarizes key information about statistical objects in tidy tibbles. This
# makes it easy to report results, create plots and consistently work with large
# numbers of models at once
library(broom)

# Create a nested dataframe
iris %>% 
  group_by(Species) %>% 
  nest() %>% 

  # Develop a separate linear regression model 
  mutate(fit = map(data, ~lm(Sepal.Length ~ Sepal.Width, data = .)), 
         results = map(fit, tidy)) %>% 
  unnest(results) %>% 
  
  # Filter & print p values
  filter(term == "Sepal.Width") %>% 
  select(Species, term, p.value)


```


